Piecewise Regression for Nutritional Outcomes Pre- and Post-CFTR Modulator

### 2022-05-18 StatNet/knot_models.qmd
Update knot_models.qmd

---
title: "Piecewise Regression for Nutritional Outcomes Pre- and Post-CFTR Modulator"
author: "Tim Vigers"
format: revealjs
editor: visual
bibliography: /Users/timvigers/Dropbox/Miscellaneous/references.bib
csl: /Users/timvigers/GitHub/styles/american-medical-association-alphabetical.csl
cache: false
---

@@

## Pre-/Post-CFTR Data

-   Limited to 3 years pre/post

## Vitamin A Plot

```{r}
vita_plot = ggplot(cf,aes(x = Time,y = Vitamin.A,color = factor(Patient.ID))) + 
  geom_point(alpha = 0.6) +
  ylab("Vitamin A") + theme_bw() + theme(legend.position = "none")
```

```{r}
mod = lmer(Vitamin.A ~ Time + c1 + c2 + (1|Patient.ID),data = cf)
```

### To Do

-   Change missing c2 values to 0

-   Write out mixed model with two knots

-   Summarize results

@@

-   The `mcp` package provides an easy interface for fitting Bayesian regression models with multiple change points.

## On the triceps data

```{r}
# Define the model
model = list(
  triceps ~ 1 + age,          # intercept + slope
  ~ 0 + age,  # joined slope with 3 change points
  ~ 0 + age,
  ~ 0 + age
)
bayes_fit = mcp(model, data = triceps, sample = "both")
plot(bayes_fit)
```

@@

@@

```{r}
s = summary(bayes_fit)
kable(s,digits = 3)
```

---

@@

```{r}
hypothesis(bayes_fit, "cp_1 = 5")
hypothesis(bayes_fit, "cp_2 = 10")
hypothesis(bayes_fit, "cp_3 = 20")
```

@@

---

```{r}
# Define the model
model = list(
  Vitamin.A ~ 1 + Time, # intercept + slope
  ~ 0 + Time  # joined slope
)
fit = mcp(model, data = cf)
plot(fit)
```

-   What does this mean? Why is the change point estimated to be before 0?

---

```{r}
# Dirichlet prior on change point
d_prior = list(
  cp_1 = "dnorm(0,1)"
)
fit_cp_prior = mcp(model, prior = d_prior,data = cf)
plot(fit_cp_prior)
```

---

>>>>

---
title: "Piecewise Regression for Nutritional Outcomes Pre- and Post-CFTR Modulator"
author: "Tim Vigers"
format: revealjs
editor: visual
bibliography: /Users/timvigers/Dropbox/Miscellaneous/references.bib
csl: /Users/timvigers/GitHub/styles/american-medical-association-alphabetical.csl
cache: true
---

@@

## Pre-/Post-CFTR Modulator Data

-   Vitamin D measures at annual visits from CF registry and our site.

-   Some patients started on ivacaftor who then transitioned to ETI.

## Pre-/Post-CFTR Modulator Data

```{r}
vita_plot = ggplot(cf,aes(x = Time,y = Vitamin.D,color = factor(Patient.ID))) + 
  geom_point() + geom_line(alpha = 0.2) +
  ylab("Vitamin D") + theme_bw() + theme(legend.position = "none")
vita_plot
```

## Pre-/Post-CFTR Modulator Model

-   The model is similar to triceps data example, but the second knot varies by patient and we include a random intercept to account for repeated measures:

$$
\text{Vitamin D}_{ij} = \beta_0 + \beta_1 \text{Time}_{ij} +\\ \beta_2(\text{Time}_{ij}-k_1)_+ + \beta_3(\text{Time}_{ij}-k_{2i})_+ + b_i + \epsilon_i
$$

## Pre-/Post-CFTR Modulator Model

$$
(x-k_1)_+=
\left\{ 
  \begin{array}{ c l }
    0, & \quad \textrm{if } \text{Time} < 0 \\
    x-k_1,                 & \quad \textrm{if } \text{Time} \geq 0 
  \end{array}
\right.
$$

and

$$
(x-k_2)_+=
\left\{ 
  \begin{array}{ c l }
    0, & \quad \textrm{if } \text{time before }2^{nd}\text{ mod}  \\
    \text{time after }2^{nd}\text{ mod},                 & \quad \textrm{if } \text{time after }2^{nd}\text{ mod}
  \end{array}
\right.
$$

and

$$
\epsilon \sim N(0,\sigma^2_{\epsilon})\text{ and } b\sim N(0,\sigma^2_b)
$$

## Pre-/Post-CFTR Modulator Results

```{r}
mod = lmer(Vitamin.D ~ Time + c1 + c2 + (1|Patient.ID),data = cf)
plot_df = mod@frame
plot_df$pred = predict(mod)
ggplot(plot_df,aes(x = Time,y = Vitamin.D,group = factor(Patient.ID))) +
    geom_line(aes(y = pred),alpha = 0.5) + geom_point(aes(y = pred),alpha = 0.5) +
    theme_bw() + theme(legend.position = "none") + xlab("Time (years)")
```

## Pre-/Post-CFTR Modulator Results

```{r}
kable(broom.mixed::tidy(mod,"fixed")[,-1],digits = 3)
```

@@

-   The `mcp` package provides an easy interface for fitting Bayesian regression models with multiple change points.
    -   Essentially it translates `lme4` syntax into JAGS.
-   Uninformative "t-tail" priors are used unless the user specifies something else.
    -   "The default priors should be suitable for estimation in the absence of strong prior knowledge."[@lindelovMcpPackageRegression2020a]

## On the Triceps Data

```{r}
#| echo: true
# Define the model
model = list(
  triceps ~ 1 + age,          # intercept + slope
  ~ 0 + age,  # joined slope 
  ~ 0 + age,  # joined slope 
  ~ 0 + age   # joined slope 
)
```

## On the Triceps Data

```{r}
#| include: false
bayes_fit = mcp(model, data = triceps, sample = "both")
s = summary(bayes_fit)
p = plot(bayes_fit)
```

```{r}
print(p)
```

@@

## On the Triceps Data

@@

```{r}
kable(s,digits = 3)
```

## On the Triceps Data

@@

```{r}
#| echo: true
hypothesis(bayes_fit, "cp_1 = 5")
hypothesis(bayes_fit, "cp_2 = 10")
hypothesis(bayes_fit, "cp_3 = 20")
```

@@

## A Bayesian Model for Pre-/Post-CFTR Modulator

-   `mcp` can model joined slopes with varying change points:

```{r}
#| echo: true
# Define the model
model = list(
  Vitamin.D ~ 1 + Time, # intercept + slope
  ~ 0 + Time,  # joined slope
  1 + (1|Patient.ID) ~ 0 + Time
)
```

## A Bayesian Model for Pre-/Post-CFTR Modulator

```{r}
#| include: false
fit = mcp(model, data = cf)
s = summary(fit)
p = plot(fit)
```

```{r}
kable(s,digits = 3)
```

## A Bayesian Model for Pre-/Post-CFTR Modulator

```{r}
print(p)
```

-   This looks terrible!

## A Bayesian Model for Pre-/Post-First CFTR Modulator

-   Try fitting on just those with one modulator, with a N(0,1) prior on a change at 0:

```{r}
#| echo: true
# Define the model
model = list(
  Vitamin.D ~ 1 + Time, # intercept + slope
  ~ 0 + Time  # joined slope
)
priors = list(
  cp_1 = "dnorm(0,1)"
)
```

```{r}
#| include: false
fit = mcp(model,prior = priors,data = cf)
s = summary(fit)
p = plot(fit)
```

## A Bayesian Model for Pre-/Post-First CFTR Modulator

```{r}
kable(s,digits = 3)
```

## A Bayesian Model for Pre-/Post-First CFTR Modulator

```{r}
print(p)
```

## Limitations

-   Only 20 two-modulator participants with vitamin D values after starting ETI (21 measures among them).

    -   Large change in slope after second modulator start is likely driven by a few influential observations.

-   Somewhat arbitrarily limited pre-modulator values to \< 5 years from modulator start.

-   General data quality issues that we are in the process of resolving.

## Questions

-   How to handle truncated data (e.g. a value of "\< 0.1")?

-   Should I model the two-modulator and one-modulator groups separately, but still use piecewise models?

-   Any general advice (particularly for Bayesian methods)?
